Dominance Regions for Imaginary $\bfg$-vectors

### 2021-09-06 dominance_regions.tex
Some progress towards proving the recursions for compatible pairs on dominant Dyck paths.

\documentclass{amsart}
\usepackage{amsmath,amssymb,latexsym,color,ulem}
\usepackage[margin=1in]{geometry}
\usepackage{tikz}
\usepackage[bookmarks=true, bookmarksopen=true, bookmarksdepth=3,bookmarksopenlevel=2, colorlinks=true, linkcolor=blue, citecolor=blue, filecolor=blue, menucolor=blue, urlcolor=blue]{hyperref}

@@

  \section{Bases From Compatible Pairs on Dyck Paths}
  \sayDR{This is a bit of a repetition of the above introduction to Dyck paths, but with slightly different goals it seemed that a modified exposition was more appropriate.}
  For $a_1,a_2\in\ZZ_{\ge0}$, choose a Dyck path $D$ in the lattice rectangle $[0,a_1]\times[0,a_2]$.
  When we wish to emphasize the upper right bounding corner we will write $D[a_1,a_2]$.
  Consider $D$ as the set of its edges taken as a totally ordered set along the natural labeling from $(0,0)$ to $(a_1,a_2)$.
  It may be convenient to identify $D$ with the ordered set $[1,a_1+a_2]$.
  Write $D=D_1\sqcup D_2$, where $D_1$ is the set of horizontal edges and $D_2$ is the set of vertical edges.
  Clearly, $|D_1|=a_1$ and $|D_2|=a_2$.

  Observe that the data of $D$ is just the data of a bipartition of a finite totally ordered set $D=D_1\sqcup D_2$.

@@

  \begin{lemma}\mbox{}
    \begin{enumerate}
      \item If $a_1 < b a_2$, then $\mu_{V,b}\big(D^{max}[a_1,a_2]\big)=D^{max}[a_2,ba_2-a_1]$.
      \item If $a_2 < c a_1$, then $\mu_{H,c}\big(D^{max}[a_1,a_2]\big)=D^{max}[ca_1-a_2,a_1]$.
    \end{enumerate}
  \end{lemma}
  \begin{proof}
    The proof is identical to the proof of \cite[Lemma 2.4]{rupel} with appropriate notation changes.
  \end{proof}

  For $e<e'\in D$, write $ee'$ for the subpath of $D$ beginning with $e$ and ending with $e'$.
  \begin{definition}
    A subset $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$ is \emph{compatible} if: for every $h\in H$ and $v\in V$ with $h<v$, there exists an edge $e\in hv$ such that one of the following conditions is satisfied
    \[e\ne v \quad \text{and} \quad |he\cap D_2|=c|he\cap D_1|\]
    \[e\ne h \quad \text{and} \quad |ev\cap D_1|=b|ev\cap D_2|\]
    Write $\cC_D$ for the collection of all compatible subsets of $D$.
  \end{definition}
  \begin{remark}
    In the construction of the greedy basis, it is necessary to consider $D$ as closed cycle by identifying $(0,0)$ with $(a_1,a_2)$ and allowing subpaths $ee'$ with $e'<e$ to ``wrap around'' $D$ in the obvious way.
    When we wish to emphasize that this condition should be imposed, we denote the Dyck path $D$ instead by $D_\WA$.
    For $(a_1,a_2)=(u_{m+1,+},u_{m,-})$ or $(u_{m,+},u_{m+1,-})$, the wrap around condition does not restrict compatibility and can thus safely be omitted.
  \end{remark}

@@

  \begin{theorem}[\cite{lee-li-zelevinsky}]
    For $(a_1,a_2)=(u_{m+1,+},u_{m,-})$ or $(u_{m,+},u_{m+1,-})$ for $m\ge0$, the element $x_{D^{max}[a_1,a_2]}$ is a non-initial cluster variable.
    Otherwise, the element $x_{D^{max}_\WA[a_1,a_2]}$ is an element of the greedy basis
  \end{theorem}

@@

    For $t>0$, define the \emph{Dyck path suffix} $\overleftarrow{D}^t$ as the maximal Dyck path with $t$ horizontal edges starting from $(0,0)$ and terminating at $(t,h)$ while staying below the line $y-h=\frac{bc-\sqrt{bc(bc-4)}}{2b}(x-t)$, where $h=h_t=\left\lceil t \frac{bc-\sqrt{bc(bc-4)}}{2b}\right\rceil$.
    Call this $h$ the \emph{height} of $t$.
  \end{definition}
  \begin{remark}
    There often exist multiple choices of $s$ with a given depth $d$ and multiple choices of $t$ with a given height $h$.
  \end{remark}

@@

  \begin{theorem}
    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_1+ba_2,-a_2)\in\cI$, we have $\mu_{V,b}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[a_2,ba_2-a_1]$ and $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[ca_1-a_2,a_1]$.
  \end{theorem}
  \begin{proof}
    We prove the result for $\mu_{H,c}$ as the result for $\mu_{V,b}$ immediately follows from $\mu_{H,b}$ being its inverse.
    Write $D'$ for the Dyck path $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)$ and observe that by construction $D'$ has $ca_1-a_2$ horizontal edges and $a_1$ vertical edges.
    Write $v'_1,\ldots,v'_{a_1}$ for the vertical edges of $D'$ with $v'_r$ immediately preceded by $\ell_r$ horizontal edges.
    We have to show (i) that every lattice point on $D'$ stays below the lines 
    \[y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x \qquad \text{and} \qquad y-a_1=\frac{bc-\sqrt{bc(bc-4)}}{2c}\big(x-(ca_1-a_2)\big)\]
    while (ii) any lattice point above $D'$ is also above one of these lines.

@@

  Apart from being the prefix and suffix for the dominant Dyck path, the Dyck paths $\overrightarrow{D}^s$ and $\overleftarrow{D}^t$ can be observed inside the maximal Dyck paths defining cluster variables as follows.
  \begin{lemma}\mbox{}
    \begin{itemize}
      \item For $s>0$ and any $m \ge 1$ so that $s < u_{m+1,-}$, the Dyck path prefix $\overrightarrow{D}^s$ is an initial segment of the maximal Dyck path $D^{max}[u_{m,+},u_{m+1,-}]$.
      \item For $t>0$ and any $m \ge 1$ so that $t < u_{m+1,+}$, the Dyck path suffix $\overleftarrow{D}^t$ is a final segment of the maximal Dyck path $D^{max}[u_{m+1,+},u_{m,-}]$.
    \end{itemize}
  \end{lemma}

@@

>>>>

\documentclass{amsart}
\usepackage{amsmath,amssymb,latexsym,color,ulem}
\usepackage[margin=1.5in]{geometry}
\usepackage{tikz}
\usepackage[bookmarks=true, bookmarksopen=true, bookmarksdepth=3,bookmarksopenlevel=2, colorlinks=true, linkcolor=blue, citecolor=blue, filecolor=blue, menucolor=blue, urlcolor=blue]{hyperref}

@@

  \section{Bases From Compatible Pairs on Dyck Paths}
  \sayDR{This is a bit of a repetition of the above introduction to Dyck paths, but with slightly different goals it seemed that a modified exposition was more appropriate.}
  For $a_1,a_2\in\ZZ_{\ge0}$, choose a Dyck path $D$ in the lattice rectangle $[0,a_1]\times[0,a_2]$.
  When we wish to emphasize the upper right bounding corner we will write $D[a_1,a_2]$.
  Consider $D$ as the set of its edges taken as a totally ordered set along the natural labeling from $(0,0)$ to $(a_1,a_2)$.
  It may be convenient to identify $D$ with the ordered set $[1,a_1+a_2]$.
  Write $D=D_1\sqcup D_2$, where $D_1=\{h_1,\ldots,h_{a_1}\}$ is the set of horizontal edges and $D_2=\{v_1,\ldots,v_{a_2}\}$ is the set of vertical edges.
  Clearly, $|D_1|=a_1$ and $|D_2|=a_2$.

  Observe that the data of $D$ is just the data of a bipartition of a finite totally ordered set $D=D_1\sqcup D_2$.
  Given a Dyck path $D=D_1\sqcup D_2$ in the lattice rectangle $[0,a_1]\times[0,a_2]$, write $\overline{D}=\overline{D}_1\sqcup \overline{D}_2$ for the Dyck path in the lattice rectangle $[0,a_2]\times[0,a_1]$ with $\overline{D}_1=a_1+a_2+1-D_2$ and $\overline{D}_2=a_1+a_2+1-D_1$.
  This operation can be visualized as transposing the Dyck path.

  Given any quantity or object $P$ whose definition depends on $b$ and $c$, write $P^\vee$ for the same quantity or object defined with $b$ and $c$ interchanged.
  In particular, we have $u^\vee_{m,\varepsilon}=u_{m,-\varepsilon}$ for any $m$ and any $\varepsilon=\pm1$.

@@

  \begin{lemma}\mbox{}
    \label{le:maximal Dyck path recursion}
    \begin{enumerate}
      \item If $a_1 < b a_2$, then $\mu_{V,b}\big(D^{max}[a_1,a_2]\big)=D^{max}[a_2,ba_2-a_1]$.
      \item If $a_2 < c a_1$, then $\mu_{H,c}\big(D^{max}[a_1,a_2]\big)=D^{max}[ca_1-a_2,a_1]$.
    \end{enumerate}
  \end{lemma}
  \begin{proof}
    The proof is identical to the proof of \cite[Lemma 2.4]{rupel} with appropriate notation changes, c.f. Theorem~\ref{th:dominant Dyck path recursion} below.
  \end{proof}

  As an immediate consequence, we determine the structure of certain maximal Dyck paths.
  We introduce the following notation for convenience:
  \[d_m:=\begin{cases} b & \text{if $m$ is even;}\\ c & \text{if $m$ is odd;}\end{cases} \qquad \text{and} \qquad d^\vee_m:=\begin{cases} c & \text{if $m$ is even;}\\ b & \text{if $m$ is odd;}\end{cases}\]
  The following was proven in \cite[Corollary 2.6]{rupel} and follows by a simple induction using Lemma~\ref{le:maximal Dyck path recursion}.
  \begin{corollary}
    Assume $b,c\ge2$ (this can be dropped with extra notation and slight modification).
    Then, for $\varepsilon=\pm1$ and $m\ge1$, the maximal Dyck path $D_{m,\varepsilon}:=D^{max}[u_{m,\varepsilon},u_{m-1,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $D_{1,\varepsilon}$ consists of a single horizontal edge;
      \item $D_{2,\varepsilon}$ consists of $d_2$ horizontal edges followed by a vertical edge;
      \item for $m\ge3$, $D_{m,\varepsilon}$ consists of $d_m-1$ copies of $D_{m-1,\varepsilon}$ followed by a copy of $D_{m-1,\varepsilon}$ with its first $D_{m-2,\varepsilon}$ removed;
    \end{itemize}
    and the maximal Dyck path $\overline{D}^\vee_{m,\varepsilon}:=D^{max}[u_{m-1,\varepsilon},u_{m,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $\overline{D}^\vee_{1,\varepsilon}$ consists of a single vertical edge;
      \item $\overline{D}^\vee_{2,\varepsilon}$ consists of a horizontal edge followed by $d^\vee_2$ vertical edges;
      \item for $m\ge3$, $\overline{D}^\vee_{m,\varepsilon}$ consists of a copy of $\overline{D}^\vee_{m-1,\varepsilon}$ with its last $\overline{D}^\vee_{m-2,\varepsilon}$ removed followed by $d^\vee_m-1$ copies of $\overline{D}^\vee_{m-1,\varepsilon}$.
    \end{itemize}
  \end{corollary}

  For $e \le e'\in D$, write $ee'$ for the subpath of $D$ beginning with $e$ and ending with $e'$.
  When $e=e'$, we take $ee'$ to be the subpath consisting of only the edge $e$.
  \begin{definition}
    For $H\sqcup V\subset D$ with $H\subset D_0$ and $V\subset D_2$ together with a subpath $ee'\subset D$, define the \emph{horizontal shadow statistic}:
    \begin{equation}
      \label{eq:horizontal shadow statistic}
      f_H(ee')=-|ee'\cap D_2|+c|ee'\cap H|
    \end{equation}
    and the \emph{vertical shadow statistic}:
    \begin{equation}
      \label{eq:vertical shadow statistic}
      f_V(ee')=-|ee'\cap D_1|+b|ee'\cap V|.
    \end{equation}
  \end{definition}
  \begin{remark}
    Observe that the shadow statistics are additive with respect to concatenation of subpaths.
  \end{remark}
  \begin{definition}
    A subset $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$ is \emph{compatible} if: for every $h\in H$ and $v\in V$ with $h<v$, there exists an edge $e\in hv$ such that one of the following conditions is satisfied
    \[\tag{HGC} e\ne v \quad \text{and} \quad f_H(he)=0\]
    \[\tag{VGC} e\ne h \quad \text{and} \quad f_V(ev)=0\]
    Write $\cC_D$ for the collection of all compatible subsets of $D$.
  \end{definition}
  \begin{remark}
    In the construction of the greedy basis, it is necessary to consider $D$ as closed cycle by identifying $(0,0)$ with $(a_1,a_2)$ and allowing subpaths $ee'$ with $e'<e$ to ``wrap around'' $D$ in the obvious way.
    When we wish to emphasize that this condition should be imposed, we denote the Dyck path $D$ instead by $D_\WA$.
    For $(a_1,a_2)=(u_{m+1,+},u_{m,-})$ or $(u_{m,+},u_{m+1,-})$, the wrap around condition does not restrict compatibility and can thus safely be omitted.
  \end{remark}

  \begin{definition}
    For $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$, define the \emph{shadow paths} as follows:
    \begin{itemize}
      \item For $h\in D_1$, write $D(h;H)$ for the shortest subpath $he \subset D$ with $f_H(he)=0$ or, if no such subpath exists, set $D(h;H)=hv_{a_2}$.
      \item For $h\in D_1$, write $D_\WA(h;H)$ for the shortest subpath $he \subset D_\WA$ with $f_H(he)=0$ or, if no such subpath exists, set $D_\WA(h;H)=D$.
      \item For $v\in D_2$, write $D(v;V)$ for the shortest subpath $ev \subset D$ with $f_V(ev)=0$ or, if no such subpath exists, set $D(v;V)=h_1 v$.
      \item For $v\in D_2$, write $D_\WA(v;V)$ for the shortest subpath $ev \subset D_\WA$ with $f_V(ev)=0$ or, if no such subpath exists, set $D_\WA(v;V)=D$.
    \end{itemize}
  \end{definition}

  \begin{lemma}
    Consider $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$.
    \begin{enumerate}
      \item For $h\in H$, take $he$ to be either the path $D(h;H)$ or $D_\WA(h;H)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $he'\subsetneq he$, we have $f_H(he')>0$;
          \item for any proper subpath $e'e\subsetneq he$, we have $f_H(e'e)<0$;
          \item $e\in V$.
        \end{enumerate}
      \item For $v\in V$, take $ev$ to be either the path $D(v;V)$ or $D_\WA(v;V)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $e'v\subsetneq ev$, we have $f_V(e'v)>0$;
          \item for any proper subpath $ee'\subsetneq ev$, we have $f_V(ee')<0$;
          \item $e\in H$.
        \end{enumerate}
    \end{enumerate}
  \end{lemma}
  \begin{proof}
    We prove (1), the proof of (2) is similar.
    To see (a), note that as $e'$ traverses the path $he$ from $h$ to $e$, the value of $f_H(he')$ either increases by $c$, stays constant, or decreases by 1.
    Since $h\in H$, we have $f_H(hh)=c>0$ and so the value of $f_H(he')$ stays positive until it first becomes 0 when $e'=e$.
    Part (b) then follows from the additivity of $f_H$ and part (c) is the case when $e'=e$ here.
  \end{proof}

@@

  \begin{theorem}[\cite{lee-li-zelevinsky}]
    For $(a_1,a_2)=(u_{m,+},u_{m-1,-})$ or $(u_{m-1,+},u_{m,-})$ for $m\ge1$, the element $x_{D^{max}[a_1,a_2]}$ is a non-initial cluster variable.
    Otherwise, the element $x_{D^{max}_\WA[a_1,a_2]}$ is an element of the greedy basis
  \end{theorem}

@@

    For $t>0$, define the \emph{Dyck path suffix} $\overleftarrow{D}^t$ as the maximal Dyck path with $t$ horizontal edges starting from $(0,0)$ and terminating at $(t,h)$ while staying below the line $y-h=\frac{bc-\sqrt{bc(bc-4)}}{2b}(x-t)$, where $h=h_t=\left\lceil t \frac{bc-\sqrt{bc(bc-4)}}{2b}\right\rceil$.
    Call this $h$ the \emph{height} of $t$.
  \end{definition}
  \begin{remark}
    There often exist multiple choices of $s$ with a given depth $d$ and multiple choices of $t$ with a given height $h$.
  \end{remark}
  \sayDR{ToDo: describe the analogue of Corollary 7.5 for the Dyck path prefixes and suffixes.}

@@

  \begin{theorem}
    \label{th:dominant Dyck path recursion}
    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_1+ba_2,-a_2)\in\cI$, we have $\mu_{V,b}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[a_2,ba_2-a_1]$ and $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[ca_1-a_2,a_1]$.
  \end{theorem}
  \begin{proof}
    We prove the result for $\mu_{H,c}$ as the result for $\mu_{V,b}$ immediately follows from $\mu_{H,b}$ being its inverse.
    Write $D'$ for the Dyck path $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)$ and observe that by construction $D'$ has $ca_1-a_2$ horizontal edges and $a_1$ vertical edges.
    Write $v'_1,\ldots,v'_{a_1}$ for the vertical edges of $D'$ with $v'_r$ immediately preceded by $\ell_r$ horizontal edges.
    We have to show (i) that every lattice point on $D'$ stays below the lines 
    \[y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x \qquad \text{and} \qquad y-a_1=\frac{bc-\sqrt{bc(bc-4)}}{2c}\big(x-(ca_1-a_2)\big)\]
    while (ii) any lattice point above $D'$ is also above one of these lines.

@@

  Apart from being the prefix and suffix for the dominant Dyck path, the Dyck paths $\overrightarrow{D}^s$ and $\overleftarrow{D}^t$ can be observed inside the maximal Dyck paths defining cluster variables as follows.
  \begin{lemma}\mbox{}
    \begin{itemize}
      \item For $s>0$ and any $m \ge 2$ so that $s < u_{m,-}$, the Dyck path prefix $\overrightarrow{D}^s$ is an initial segment of the maximal Dyck path $D^{max}[u_{m-1,+},u_{m,-}]$.
      \item For $t>0$ and any $m \ge 2$ so that $t < u_{m,+}$, the Dyck path suffix $\overleftarrow{D}^t$ is a final segment of the maximal Dyck path $D^{max}[u_{m,+},u_{m-1,-}]$.
    \end{itemize}
  \end{lemma}

  This can be made more precise as follows.
  \begin{lemma}
    For $s>0$, choose $m\ge2$ minimal so that $s < u_{m,-}$
  \end{lemma}

@@

  \begin{definition}
    Suppose $D^{dom}[a_1,a_2]$ decomposes as $\overrightarrow{D}^s \sqcup \overleftarrow{D}^t$.
    Call a pair of subsets $H \sqcup V$ \emph{piecewise compatible} if the intersections with $\overrightarrow{D}^s$ and $\overleftarrow{D}^t$ are compatible on the subpaths.
  \end{definition}

  \begin{theorem}
    Suppose $H \sqcup V$ is a piecewise compatible subset of $D^{dom}[a_1,a_2]$.
    \begin{itemize}
      \item If the last vertical edge of $\overrightarrow{D}^s$ is contained in $V$, then $H \sqcup V$ is compatible.
      \item If the first horizontal edge of $\overleftarrow{D}^t$ is contained in $H$, then $H \sqcup V$ is compatible.
    \end{itemize}
  \end{theorem}
